---
title: "Assignment 5: Data Visualization"
author: "Rachel Williams"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse);library(lubridate);library(here);library(cowplot)
here()
PeterPaul_processed <- read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"))
NiwotRidge_litter <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"))
                                
#2 
class(PeterPaul_processed$sampledate)
PeterPaul_processed$sampledate <- ymd(PeterPaul_processed$sampledate)
class(PeterPaul_processed$sampledate)
class(NiwotRidge_litter$collectDate)
NiwotRidge_litter$collectDate <- ymd(NiwotRidge_litter$collectDate)
class(NiwotRidge_litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
mytheme <-theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        plot.background = element_rect(color="gray"), plot.title=element_text(size=16, face="bold", hjust=.5))

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
PhosGraph <- PeterPaul_processed %>%
ggplot(
    aes(
      x=tp_ug,
      y=po4),
    ) + geom_point() + ggtitle("Phosphorus vs. Phosphate")  + ylim(0, 60) + xlab("Total Phosphorus") + ylab("Total Phosphate")  + geom_smooth(method = "lm", se = FALSE) +
  mytheme 
PhosGraph
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
PeterPaul_processed$month <-factor(PeterPaul_processed$month,
       levels= 1:12,
       labels = month.abb)

tempbox <- PeterPaul_processed %>%
  ggplot(aes(x=month, y=temperature_C, fill=lakename)) +
  geom_boxplot() 
tempbox

TPbox <- PeterPaul_processed %>%
  ggplot(aes(x= month, y= tp_ug, fill=lakename)) +
  geom_boxplot()
TPbox

TNbox <- PeterPaul_processed %>%
  ggplot(aes(x=month, y=tn_ug, fill=lakename)) +
  geom_boxplot()
TNbox

combined_plot <- plot_grid(tempbox, TPbox, TNbox, ncol=2)
combined_plot
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Typically the variables of interest increase during the summer with a peak in July ot August. Peter Lake seems to have greater amounts of TP and TN than Paul lake. Both lakes have similar temperatures to each other each month. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

needles <- NiwotRidge_litter[NiwotRidge_litter$functionalGroup == "Needles",]

Needlemass <- needles %>%
ggplot(aes(x=collectDate, y=dryMass, color = nlcdClass)) + geom_point()
Needlemass

#7

Needlemass2 <- needles %>%
  ggplot(aes(x=collectDate, y=dryMass)) + geom_point() +
  facet_wrap(facets= vars(nlcdClass),ncol = 3, nrow = 1)
Needlemass2

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective because plot 6 has overlapping points that make it hard to read. In plot 7 you can look at each location seperately and compare them if needed. If plot 7 had differentiated colors between all three areas it would also be an effective way to visualize the difference without covering other points.   